iSEE: Interactive SummarizedExperiment Explorer

### 2018-05-05 manuscript/main.tex
Some tweaks to the text

\section*{Introduction}
Interactive data exploration is critical to the analysis and comprehension of the data generated by high-throughput biological assays such as genomics.
Exploration drives the formation of novel data-driven hypotheses prior to a more rigorous statistical analysis, and enables diagnosis of potential problems such as batch effects and low-quality samples.
To this end, visualization of the data in an intuitive and interactive interface is crucial, allowing researchers to examine the data from different perspectives across samples (e.g., experimental replicates, patients, single cells) and features (e.g., genes, transcripts, proteins, genomic regions).\\

@@

\section*{Taking a SummarizedExperiment input}
The iSEE interface is initialized with a single function call accepting a SummarizedExperiment object as input \citep{huber2015orchestrating}.
Each instance of the SummarizedExperiment class stores one or more matrices of experimental observations as ``assays'', where rows and columns represent genomic features and biological samples, respectively.
For instance, individual assays may represent gene expression matrices of raw counts along with different normalization strategies.
In addition, per-feature or per-sample variables are stored in the ``rowData'' and ``colData'', respectively; those may include experimental metadata as well as any analysis results.\\

@@

\begin{itemize}
\item Column data plots, for visualizing sample metadata stored in the colData slot of SummarizedExperiment object.
\item Feature assay plots, for visualizing expression values for a particular feature (e.g., gene) across samples from any assay in the SummarizedExperiment object.
Despite its name, this plot type is applicable to data modalities beyond gene expression.
\item Row statistics tables, to present the contents of the rowData slot for the SummarizedExperiment object.
\item Row data plots, for visualizing sample metadata stored in the rowData slot of a SummarizedExperiment object.
\item Heat maps, to visualize assay data for multiple features where samples are ordered by one or more colData fields.
\item Reduced dimension plots, which display any two dimensions from pre-computed dimensionality reduction results (e.g., from PCA or $t$-SNE).
These results are taken from the reducedDim slot of any SingleCellExperiment object supplied to iSEE.
\end{itemize}

\begin{figure*}[t]
\includegraphics[width=\textwidth]{pics/Fig1.pdf}
\caption{iSEE uses a customisable multi-panel layout (A) that simultaneously displays one or more panels of various types, where each panel type visualizes a different aspect of the data.
New panels of any type can be added (i), and all panels can be removed, reordered or resized (ii).
Panel types are available to visualize sample-based reduced dimensionality embeddings (iii), sample-level metadata (iv), and experimental observations across samples for each feature (v).
Other panel types include row statistics tables (vi), to facilitate searching across features and their metadata; heatmaps (vii), to visualize experimental observations for multiple features; and feature-level metadata plots.
Panels of each type are colour-coded for ease of interpretation.
(B) Information can be transmitted between panels according to a user-specified scheme.
Here, the selection of feature $X$ in the row statistics table determines the y-axis of the feature assay plot, and colours the samples in the reduced dimension plot by the expression of $X$.
Selection of points in the reduced dimension plot (dotted blue line) also determines the samples that are shown in the column data (i.e., sample metadata) plot;
further selection of points in the column data plot determines the samples that are shown in the heatmap.
}
\label{fig:iSEE}
\end{figure*}

Each sample is represented as a point in column data, feature assay and reduced dimension plots.
Similarly, each feature is represented by a point in row data plots.
For these panel types, a scatter plot is automatically produced if the selected variables on the x- and y-axes are both continuous.
If exactly one variable is categorical, points are grouped by the categorical levels and a (horizontal) violin plot is produced with points scattered within each violin.
If both variables are categorical, a ``rectangle plot'' is produced where each combination of categorical levels is represented by a rectangle with area proportional to the frequency of that combination.
Points are scattered randomly within each rectangle.
For ease of interpretation, the rectangle plot collapses to a mirrored bar plot when one of the categorical variables only has one level.

\subsection*{Custom panel colouring}
Sample-based points can be coloured according to the values of any sample-level metadata field in the colData or by the assay values of a selected feature.
Similarly, feature-based points can be coloured according to any feature-level metadata field in the rowData.
Heatmaps are coloured according to the expression values of the selected features in the chosen assay,
with additional colouring of each of the colData fields used to order the samples.
In all cases, the variable to use for colouring can be dynamically selected for each plot.
This enables users to easily examine relationships between different variables in a single plot. \\

@@

\subsection*{Dynamic linking between panels}
A key feature of iSEE is the ability to dynamically transmit information between panels (Fig.~\ref{fig:iSEE}B).
Users can define and reorganize arbitrary links between ``transmitting'' and ``receiving'' panels, whereby selections in transmitting panels control the inclusion and appearance of the corresponding data points in receiving panels.
This feature facilitates exploration of the relationships between different aspects of the data.
For example, users can easily determine co-expression patterns of genes in a particular region of a reduced dimensionality embedding -- this is achieved by selecting points in a reduced dimension plot (using the standard rectangular brush or a lasso) and transmitting that selection to any number of feature assay plots.\\

This linking paradigm extends to multiple panels, whereby a panel can transmit to multiple receivers, and a receiving panel can transmit its own selection to another plot.
Chains of linked plots allow users to mimic the arbitrarily complex gating strategies often found in analyses of flow cytometry data \citep{finak2014opencyto}.
With iSEE, this concept is extended to any assay data, feature-level or sample-level metadata present in a SummarizedExperiment object, providing a powerful framework for interrogating multiple interactions between data aspects.
Row statistics tables can also transmit to various plot types, by selecting a table row to control the coloring of sample-based points;
or by defining a subset of features to visualize in a heatmap.
Furthermore, row data plots can transmit to row statistics tables, whereby selection of points in the former will subset the latter.

@@

iSEE automatically memorizes the exact R code that was used to generate every plot, extending previous work by \cite{marini2016interrepro}.
This code is fully accessible to users at any time during the run-time of the interface.
By integrating the code reported by iSEE into their own scripts, users can easily reproduce the results of any exploratory analysis.
Similarly, the code required to reproduce the current state of the interface can also be reported.
This can be used in startup scripts to launch an iSEE instance in any preferred layout, including the panel organization, variable selection, colouring schemes, links between panels and even individual brushes and lasso selections.

\subsection*{Additional functionalities}
Row statistics tables can be augmented with dynamic annotation based on the selected row, linking to online resources such as Ensembl or Entrez. % do we need to cite them? at least ENSEMBL has updated citable papers
For large data sets, points can be downsampled in a density-dependent manner to accelerate rendering of the plots, improving the responsiveness of the interface without compromising the fidelity of the visualization.
Users can also include a bespoke step-by-step ``tour'' of their data set via the rintrojs package \cite{ganz2016rintrojs}, guiding the audience through an examination of the salient features in the data.

@@

\section*{Conclusion}
iSEE provides a general interactive interface for visual exploration of high-throughput biological data sets.
Any study that can be represented in a SummarizedExperiment object can be used as input, allowing iSEE to accommodate a diverse range of `-omics data sets.
The interface is flexible and can be dynamically customized by the user; supports exploration of interactions between data aspects through colouring and linking between panels;
and provides transparency and reproducibility during the interactive analysis, through code tracking and state reporting. 
The most obvious use of iSEE is that of data exploration for hypothesis generation during the course of a research project.
However, we also anticipate that public instances of iSEE will accompany publications to enable authors to showcase salient aspects of their work using guided tours.

>>>>

\section*{Introduction}
Interactive data exploration is critical to the analysis and comprehension of the data generated by high-throughput biological assays commonly used in genomics.
Exploration drives the formation of novel data-driven hypotheses prior to a more rigorous statistical analysis, and enables diagnosis of potential problems such as batch effects and low-quality samples.
To this end, visualization of the data in an intuitive and interactive interface is crucial for allowing researchers to examine the data from different perspectives across samples (e.g., experimental replicates, patients, single cells) and features (e.g., genes, transcripts, proteins, genomic regions).\\

@@

\section*{Taking a SummarizedExperiment input}
The iSEE interface is initialized with a single function call accepting a SummarizedExperiment object as input \citep{huber2015orchestrating}.
Each instance of the SummarizedExperiment class stores one or more matrices of experimental observations as ``assays'', where rows and columns represent genomic features and biological samples, respectively.
For instance, individual assays may represent gene expression matrices, either in the form of raw counts or as values obtained by different normalization strategies.
In addition, per-feature or per-sample variables are stored in the ``rowData'' and ``colData'', respectively; those may include experimental metadata as well as any analysis results.\\

@@

\begin{itemize}
\item Column data plots, for visualizing sample metadata stored in the colData slot of the SummarizedExperiment object.
\item Feature assay plots, for visualizing expression values for a particular feature (e.g., gene) across samples from any assay in the SummarizedExperiment object.
Despite its name, this plot type is applicable to data modalities beyond gene expression.
\item Row statistics tables, to present the contents of the rowData slot for the SummarizedExperiment object.
\item Row data plots, for visualizing feature metadata stored in the rowData slot of the SummarizedExperiment object.
\item Heat maps, to visualize assay data for multiple features where samples are ordered by one or more colData fields.
\item Reduced dimension plots, which display any two dimensions from pre-computed dimensionality reduction results (e.g., from PCA or $t$-SNE).
These results are taken from the reducedDim slot if this is available in the object supplied to iSEE.
\end{itemize}

\begin{figure*}[t]
\includegraphics[width=\textwidth]{pics/Fig1.pdf}
\caption{iSEE uses a customisable multi-panel layout (A) that simultaneously displays one or more panels of various types, where each panel type visualizes a different aspect of the data.
New panels of any type can be added (i), and all panels can be removed, reordered or resized (ii).
Panel types are available to visualize sample-based reduced dimensionality embeddings (iii), sample-level metadata (iv), and experimental observations across samples for each feature (v).
Other panel types include row statistics tables (vi), to facilitate searching across features and their metadata; heat maps (vii), to visualize experimental observations for multiple features; and feature-level metadata plots.
Panels of each type are colour-coded for ease of interpretation.
(B) Information can be transmitted between panels according to a user-specified scheme.
Here, the selection of feature $X$ in the row statistics table determines the y-axis of the feature assay plot, and colours the samples in the reduced dimension plot by the expression of $X$.
Selection of points in the reduced dimension plot (dotted blue line) also determines the samples that are shown in the column data (i.e., sample metadata) plot;
further selection of points in the column data plot determines the samples that are shown in the heat map.
}
\label{fig:iSEE}
\end{figure*}

Each sample is represented as a point in column data, feature assay and reduced dimension plots.
Similarly, each feature is represented by a point in row data plots.
For these panel types, a scatter plot is automatically produced if the selected variables on the x- and y-axes are both continuous.
If exactly one variable is categorical, points are grouped by the categorical levels and a (vertical or horizontal) violin plot is produced with points scattered within each violin.
If both variables are categorical, a ``rectangle plot'' is produced where each combination of categorical levels is represented by a rectangle with area proportional to the frequency of that combination.
Points are scattered randomly within each rectangle.
For ease of interpretation, the rectangle plot collapses to a mirrored bar plot when one of the categorical variables only has one level.

\subsection*{Custom panel colouring}
Sample-based points can be coloured according to the values of any sample-level metadata field in the colData slot or by the assay values of a selected feature.
Similarly, feature-based points can be coloured according to any feature-level metadata field in the rowData slot.
Heat maps are coloured according to the expression values of the selected features in the chosen assay,
with additional colour annotation by each of the colData fields used to order the samples.
In all cases, the variable to use for colouring can be dynamically selected for each plot.
This enables users to easily examine relationships between different variables in a single plot. \\

@@

\subsection*{Dynamic linking between panels}
A key feature of iSEE is the ability to dynamically transmit information between panels (Fig.~\ref{fig:iSEE}B).
Users can define and reorganize arbitrary links between ``transmitting'' and ``receiving'' panels, whereby selections in transmitting panels control the inclusion and appearance of the corresponding data points in receiving panels.
This feature facilitates exploration of the relationships between different aspects of the data.
For example, users can easily determine co-expression patterns of genes in a particular region of a reduced dimensionality embedding -- this is achieved by selecting points in a reduced dimension plot (using the standard rectangular brush or a lasso selection) and transmitting that selection to any number of feature assay plots.\\

This linking paradigm extends to multiple panels, whereby a panel can transmit to multiple receivers, and a receiving panel can transmit its own selection to another plot.
Chains of linked plots allow users to mimic the arbitrarily complex gating strategies often found in analyses of flow cytometry data \citep{finak2014opencyto}.
With iSEE, this concept is extended to any assay data, feature-level or sample-level metadata present in a SummarizedExperiment object, providing a powerful framework for interrogating multiple interactions between data aspects.
Row statistics tables can also transmit to various plot types, by selecting a table row to control the coloring of sample-based points;
or by defining a subset of features to visualize in a heat map.
Furthermore, row data plots can transmit to row statistics tables, whereby selection of points in the former will subset the latter.

@@

iSEE automatically memorizes the exact R code that was used to generate every plot, extending previous work by \citet{marini2016interrepro}.
This code is fully accessible to users at any time during the run-time of the interface.
By integrating the code reported by iSEE into their own scripts, users can easily reproduce the results of any exploratory analysis.
Similarly, the code required to reproduce the current state of the interface can also be reported.
This can be used in startup scripts to launch an iSEE instance in any preferred layout, including the panel organization, variable selection, colouring schemes, links between panels and even individual brushes and lasso selections.

\subsection*{Additional functionalities}
Row statistics tables can be augmented with dynamic annotation based on the selected row, linking to online resources such as Ensembl \citep{zerbino2018ensembl} or Entrez \citep{ncbi2017entrez}. % do we need to cite them? at least ENSEMBL has updated citable papers
For large data sets, points can be downsampled in a density-dependent manner to accelerate rendering of the plots, improving the responsiveness of the interface without compromising the fidelity of the visualization.
Users can also include a bespoke step-by-step ``tour'' of their data set via the rintrojs package \cite{ganz2016rintrojs}, guiding the audience through an examination of the salient features in the data.

@@

\section*{Conclusion}
iSEE provides a general interactive interface for visual exploration of high-throughput biological data sets.
Any study that can be represented in a SummarizedExperiment object can be used as input, allowing iSEE to accommodate a diverse range of `omics data sets.
The interface is flexible and can be dynamically customized by the user; supports exploration of interactions between data aspects through colouring and linking between panels;
and provides transparency and reproducibility during the interactive analysis, through code tracking and state reporting. 
The most obvious use of iSEE is that of data exploration for hypothesis generation during the course of a research project.
However, we also anticipate that public instances of iSEE will accompany publications to enable authors to showcase salient aspects of their work using guided tours.
